Hydra Battles and Cie

### 2021-05-21 doc/chapter-primrec.tex
corrections to the pdf

\begin{description}
  \item[zero] the constant function of value $0$ is primitive recursive.
\item[S] The successor function $S:\mathbb{N}\rightarrow\mathbb{N}$ is primitive recursive.
 \item[projections] For any pair $0< i\leq n$, the projection $\pi_{i,n}: \mathbb{N}^n\rightarrow\mathbb{N}$, defined by $\pi_{i,n}(x_1,x_2,\dots,x_{n})=x_i$, is primitive recursive.
\item[composition] For any $n$ and $m$, if $h: \mathbb{N}^m\rightarrow\mathbb{N}$, and
$g_0,\dots, g_{m-1}$ are primitive recursive of $n$ arguments, then the function which maps any
tuple $(x_0,\dots,x_{n-1})$ to $h(g_0(x0,\dots,x_{n-1}),\dots, g_{m-1}(x0,\dots,x_{n-1}))$ is primitive recursive.
\item[primitive recursion]
If $g: \mathbb{N}^n\rightarrow\mathbb{N}$ and $h: \mathbb{N}^{n+2}\rightarrow\mathbb{N}$ are primitive recursive, then the function from $\mathbb{N}^{n+1}$ into $\mathbb{N}$ defined by
\begin{align}
f(0,x_1,\dots,x_n)&=g(x_1,\dots,x_n)\\
f(S(p),x_1,\dots,x_n)&=h(p,f(p, x_1,\dots,x_n),  x_1,\dots,x_n)
\end{align} 
is primitive recursive.
\end{description}

@@

O'Connor's library on Gödel's incompleteness theorems contains a little more 
than 45K lines of scripts. The part dedicated to primitive recursive functions and Peano arithmetics is 32K lines long and is originally structured in 38 modules.
Thus, we propose a slow exploration of this library, through examples and exercises. Our additions to the original library 
are stored in the directory \texttt{theories/ordinals/MoreAck}.
In particular, the library~\href{../theories/html/hydras.MoreAck.AckNotPR.html}{MoreAck.AckNotPR} contains the well-known  proof that the Ackermann function is not primitive recursive (see Section~\vref{sect:ack-not-PR}).

@@

Library \texttt{primRec} allows us to consider primitive functions on \texttt{nat}, with any number of arguments, in 
curried form. This is made possible in 
\href{../theories/html/hydras.Ackermann.extEqualNat.html}{Ackermann.extEqualNat} by the following definition:

@@

\vspace{4pt}
\noindent
\emph{From\href{../theories/html/hydras.MoreAck.PrimRecExamples.html}{MoreAck.PrimRecExamples}}.
\begin{Coqsrc}
Require Import primRec.
Import extEqualNat.

@@

\begin{remark}
\label{projFunc-order-of-args}
Beware of the conventions used in the \texttt{primRec} library!
The constructor (\texttt{projFunc $n$ $m$})  is associated with the projection $\pi_{n-m,n}$ and \emph{not}
$\pi_{n, m}$.
For instance, the projection $\pi_{2,5}$ defined by $\pi_{2,5}(a,b,c,d,e)=b$ corresponds to the term
(\texttt{projFunc 5 3 H}), where \texttt{H} is a proof of $3<5$.
 This fact is reported in the comments of \texttt{primRec.v}. We presume that this convention makes it easier to define the evaluation function \texttt{evalProjFunc $n$} (see the next sub-section). Trying the other convention is left as an exercise.
\end{remark}

\subsection{A little bit of semantics} 
Please note that inhabitants of type (\texttt{PrimRec $n$}) are not \coq{} functions like \texttt{Nat.mul}, or factorial, etc. The data-type (\texttt{PrimRec $n$}) is indeed an abstract syntax for the language of primitive recursive functions. The bridge between this language and the word of usual functions
is an interpretation function (\texttt{evalprimRec $n$})  of type
$\texttt{PrimRec}\,n \rightarrow  \texttt{naryFunc}\,n$.
This function is definied by mutual recursion,  together with the  function 
(\texttt{evalprimRecS $n$ $m$}) of type 
$\texttt{PrimRecs}\,n\,m \rightarrow  \texttt{Vector.t}\,(\texttt{naryFunc}\,n)\,m$.

@@

Another example?
Let us consider the following term\footnote{Of course, we never typed this term \emph{verbatim}; we obtained it as the result of some computation we leave it the reader to guess and reproduce.}:

@@

\section{Proving that a given arithmetic function is primitive recursive}

The example in the preceding section clearly shows that, in order to prove that a given arithmetic function
(defined in \gallina{} as usual) is primitive recursive, trying to give  by hand a term  of type \texttt{PrimRec $n$} is not a good method, since such terms may be huge and complex, even for simple arithmetic functions. The method proposed in Library \texttt{primRec} is the following one:

@@

If it does not cause compatibility problems (with 
\href{https://github.com/coq-community/goedel}{goedel library} fr instance), we plan to make all theses lemmas transparent.
\end{remark}

@@

\textbf{Hint:}  You may use as a helper the function which computes the pair 
$(\texttt{fib}(n+1),\texttt{fib}(n))$. 
Library \href{../theories/html/hydras.Ackermann.cPair.html}{Ackermann.cPair} contains
the definition of the encoding of $\mathbb{N}^2$ into $\mathbb{N}$, and the proofs that 
the associated constructor and projections are primitive recursive.

\emph{You may start this exercise with the file
\url{../exercises/primrec/FibonacciPR.v}.}

@@

A possible workaround is to make \\texttt{m} be the 
decreasing argument, and define --- within \texttt{m}'s scope --- a local helper function which computes \texttt{Ack m n} for any \texttt{n}.
This way, both functions \texttt{Ack} and \texttt{Ackm} have a (structurally) strictly decreasing argument.

@@

\begin{exercise}
The file \href{../theories/html/hydras.MoreAck.Ack.html}{MoreAck.Ack} presents two other definitions\footnote{post on \href{https://stackoverflow.com/questions/10292421/error-in-defining-ackermann-in-coq} by Anton Trunov.} of the Ackermann functions based on the lexicographic ordering on $\mathbb{N}\times\mathbb{N}$.
Prove that all the four functions are extensionnally equal.
\end{exercise}

@@

\begin{Coqsrc}
Lemma nested_Ack_bound : forall k m n, 
    Ack k (Ack m n) <= Ack (2 + max k m) n.
\end{Coqsrc}

@@

First, we prove that any primitive function of two arguments is majorized.
If we look at the inductive definition of primitive recursive functions, page~\pageref{def:Primrec}, it is obvious that a proof by induction on the construction of primitive recursive functions must consider functions of any arity.

\noindent
\emph{From \href{../theories/html/hydras.Ackermann.primRec.html}{Ackermann.primRec}.}

@@

* projections *)
(forall (n m : nat) (l : m < n), P n (projFunc n m l)) ->

@@

First, we extend \texttt{max} to vectors of natural numbers (using the notations of module \texttt{VectorNotations} and some more definitions from 
\href{../theories/html/hydras.Prelude.MoreVectors.html}{Prelude.MoreVectors}). So, \texttt{t\,$A$\,$n$} is the type of vectors of $n$ elements of type $A$, and the constants \texttt{cons}, \texttt{nil}, \texttt{map}, etc., refer to vectors and not to lists. Likewise, the notation \texttt{x::v} is an abbreviation for
\texttt{VectorDef.cons x \_ v}.

@@

We have also to convert any application
$(f\,x_1\,x_2\,\dots\,x_n)$ into an application of a function 
to a single argument: the vector of all the $x_i$s.
This is already done in 
Library~\href{../theories/html/hydras.Ackermann.primRec.html}{Ackermann.primRec}.

@@

Now, it remains to prove that any primitive function is majorized by \texttt{Ack}.
The three base cases of construction of primitive recursive function are as follows:

@@

\subsection{The length of standard hydra battles}

The module \href{../theories/html/hydras.Hydra.Hydra_Theorems.html}{Hydra\_Theorems} contains a proof that the function which computes the lenght of standard hydra battles is not primitive recursive. This proof is  a little more complex than the preceding one.

\subsubsection{Definitions}

@@

\subsubsection{Proof steps}

@@

Now, let us recall the following lemmas :

\begin{enumerate}
\item 
\emph{From \href{../theories/html/hydras.Epsilon0.F_alpha.html}{Epsilon0.F\_alpha}}

\begin{Coqsrc}
H_F : forall (alpha : E0) (n : nat), 
     F_ alpha (S n) <= H_ (Phi0 alpha) (S n) 
\end{Coqsrc}

\emph{Please note that our proof of this lemma is not trivial at all. It uses some lemmas of Chapters~\vref{chap:ketonen} and \vref{chap:alpha-large}.}

\item
\noindent
\emph{From \href{../theories/html/hydras.Epsilon0.L_alpha.html}{Epsilon0.L\_alpha}}

\begin{Coqsrc}
H_L_ : forall (alpha : E0) (i : nat), H_ alpha i <= L_ alpha (S i)
\end{Coqsrc}

\item

@@

\end{enumerate}

>>>>

\begin{description}
  \item[zero] the constant function of value $0$ is primitive recursive.
\item[S] The successor function $S:\mathbb{N}\rightarrow\mathbb{N}$ is primitive recursive.
 \item[projections] For any pair $0< i\leq n$, the projection $\pi_{i,n}: \mathbb{N}^n\rightarrow\mathbb{N}$, defined by $\pi_{i,n}(x_1,x_2,\dots,x_{n})=x_i$, is primitive recursive.
\item[composition] For any $n$ and $m$, if $h: \mathbb{N}^m\rightarrow\mathbb{N}$, and
$g_0,\dots, g_{m-1}: \mathbb{N}^n\rightarrow\mathbb{N}$ are primitive recursive of $n$ arguments, then the function which maps any
tuple $(x_0,\dots,x_{n-1})$ to $h(g_0(x0,\dots,x_{n-1}),\dots, g_{m-1}(x0,\dots,x_{n-1})): \mathbb{N}^n\rightarrow\mathbb{N}$ is primitive recursive.
\item[primitive recursion]
If $g: \mathbb{N}^n\rightarrow\mathbb{N}$ and $h: \mathbb{N}^{n+2}\rightarrow\mathbb{N}$ are primitive recursive, then the function from $\mathbb{N}^{n+1}$ into $\mathbb{N}$ defined by
\begin{align}
f(0,x_1,\dots,x_n)&=g(x_1,\dots,x_n)\\
f(S(p),x_1,\dots,x_n)&=h(p,f(p, x_1,\dots,x_n),  x_1,\dots,x_n)
\end{align} 
is primitive recursive.
\end{description}

@@

O'Connor's library on Gödel's incompleteness theorems contains a little more 
than 45K lines of scripts. The part dedicated to primitive recursive functions and Peano arithmetics is 32K lines long and is originally structured in 38 modules.
Thus, we propose a partial exploration of this library, through examples and exercises. Our additions to the original library --- mainly examples and counter-examples ---,
are stored in the directory \texttt{theories/ordinals/MoreAck}.

In particular, the library \href{../theories/html/hydras.MoreAck.AckNotPR.html}{MoreAck.AckNotPR} contains the well-known  proof that the Ackermann function is not primitive recursive (see Section~\vref{sect:ack-not-PR}).
MoreOver, the library \href{../theories/html/hydras.Hydra.Hydra_Theorems.html}{Hydra.Hydra\_Theorems} contains 
a proof that the length of an hydra battle (according to the initial replication factor) is not primitive recursive in general.

@@

The  \texttt{primRec} library allows us to consider primitive functions on \texttt{nat}, with any number of arguments, in 
curried form. This is made possible in 
\href{../theories/html/hydras.Ackermann.extEqualNat.html}{Ackermann.extEqualNat} by the following definition:

@@

\vspace{4pt}
\noindent
\emph{From \href{../theories/html/hydras.MoreAck.PrimRecExamples.html}{MoreAck.PrimRecExamples}}.
\begin{Coqsrc}
Require Import primRec.
Import extEqualNat.

@@

\begin{remark}
\label{projFunc-order-of-args}
Beware of the conventions used in the \texttt{primRec} library!
The constructor (\texttt{projFunc $n$ $m$})  is associated with the projection $\pi_{n-m,n}$ and \emph{not}
$\pi_{n, m}$.
For instance, the projection $\pi_{2,5}$ defined by $\pi_{2,5}(a,b,c,d,e)=b$ corresponds to the term
(\texttt{projFunc 5 3 H}), where \texttt{H} is a proof of $3<5$.
 This fact is reported in the comments of \texttt{primRec.v}. We presume that this convention makes it easier to define the evaluation function (\texttt{evalProjFunc $n$}) (see the next sub-section). Trying the other convention is left as an exercise.
\end{remark}

\subsection{A little bit of semantics} 
Please note that inhabitants of type (\texttt{PrimRec $n$}) are not \coq{} functions like \texttt{Nat.mul}, or factorial, etc. The data-type (\texttt{PrimRec $n$}) is indeed an abstract syntax for the language of primitive recursive functions. The bridge between this language and the word of usual functions
is an interpretation function (\texttt{evalprimRec $n$})  of type
$\texttt{PrimRec}\,n \rightarrow  \texttt{naryFunc}\,n$.
This function is defined by mutual recursion,  together with the  function 
(\texttt{evalprimRecS $n$ $m$}) of type 
$\texttt{PrimRecs}\,n\,m \rightarrow  \texttt{Vector.t}\,(\texttt{naryFunc}\,n)\,m$.

@@

Another example?
Let us consider the following term\footnote{Of course, we never typed this term \emph{verbatim}; we obtained it by an interactive proof the reader will be able to make after 
reading Sect.\vref{sect:proofs-of-isPR}.}:

@@

\section{Proving that a given arithmetic function is primitive recursive}
\label{sect:proofs-of-isPR}

The example in the preceding section clearly shows that, in order to prove that a given arithmetic function
(defined in \gallina{} as usual) is primitive recursive, trying to give  by hand a term  of type (\texttt{PrimRec $n$}) is not a good method, since such terms may be huge and complex, even for simple arithmetic functions. The method proposed in Library \texttt{primRec} is the following one:

@@

If it does not cause compatibility problems (with 
\href{https://github.com/coq-community/goedel}{goedel library} for instance), we plan to make all theses lemmas transparent.
\end{remark}

@@

\textbf{Hint:}  You may use as a helper the function which computes the pair \linebreak
$(\texttt{fib}(n+1),\texttt{fib}(n))$. 
Library \href{../theories/html/hydras.Ackermann.cPair.html}{Ackermann.cPair} contains
the definition of the encoding of $\mathbb{N}^2$ into $\mathbb{N}$, and the proofs that 
the associated constructor and projections are primitive recursive.
\emph{You may start this exercise with the file
\url{../exercises/primrec/FibonacciPR.v}.}

@@

A possible workaround is to make \texttt{m} be the 
decreasing argument, and define --- within \texttt{m}'s scope --- a local helper function which computes (\texttt{Ack m n}) for any \texttt{n}.
This way, both functions \texttt{Ack} and \texttt{Ackm} have a (structurally) strictly decreasing argument.

@@

\begin{exercise}
The file \href{../theories/html/hydras.MoreAck.Ack.html}{MoreAck.Ack} presents two other definitions\footnote{post on \href{https://stackoverflow.com/questions/10292421/error-in-defining-ackermann-in-coq} by Anton Trunov.} of the Ackermann functions based on the lexicographic ordering on $\mathbb{N}\times\mathbb{N}$.
Prove that the four functions are extensionnally equal.
\end{exercise}

@@

\noindent
\emph{From \href{../theories/html/hydras.MoreAck.Ack.html}{MoreAck.Ack}}.
\begin{Coqsrc}
Lemma nested_Ack_bound : forall k m n, 
    Ack k (Ack m n) <= Ack (2 + max k m) n.
\end{Coqsrc}

@@

First, we prove that any primitive function of two arguments is majorized by \texttt{Ack}.
If we look at the inductive definition of primitive recursive functions, page~\pageref{def:Primrec}, it is obvious that a proof by induction on the construction of primitive recursive functions must consider functions of any arity.

The following scheme allows us to write proofs by induction on the class of primitive recursive functions.

\vspace{4pt}
\noindent
\emph{From \href{../theories/html/hydras.Ackermann.primRec.html}{Ackermann.primRec}.}

@@

(* projections *)
(forall (n m : nat) (l : m < n), P n (projFunc n m l)) ->

@@

First, we extend \texttt{max} to vectors of natural numbers (using the notations of module \texttt{VectorNotations} and some more definitions from 
\href{../theories/html/hydras.Prelude.MoreVectors.html}{Prelude.MoreVectors}). So, (\texttt{t\,$A$\,$n$}) is the type of vectors of $n$ elements of type $A$, and the constants \texttt{cons}, \texttt{nil}, \texttt{map}, etc., refer to vectors and not to lists. Likewise, the notation \texttt{x::v} is an abbreviation for
\texttt{VectorDef.cons x \_ v}.

@@

We have also to convert any application
$(f\,x_1\,x_2\,\dots\,x_n)$ into an application of a function 
to a single argument: the vector of all the $x_i$\,s.
This is already defined in 
Library~\href{../theories/html/hydras.Ackermann.primRec.html}{Ackermann.primRec}.

@@

Now, it remains to prove that any primitive function is majorized by \texttt{Ack}.
The three base cases  are as follows:

@@

\section{The length of standard hydra battles}

The module \href{../theories/html/hydras.Hydra.Hydra_Theorems.html}{Hydra\_Theorems} contains a proof that the function which computes the length of standard hydra battles is not primitive recursive. More precisely, we consider, for a given hydra $h=\iota(\alpha)$, the length of a standard battle which starts with the replication factor $k$ (see Sect~\vref{def:L-alpha}).

This proof is  a little more complex than the preceding one.

\subsection{Definitions}

@@

\subsection{Proof steps}

@@

\subsubsection{Comparison between $F$ and $H'$}

In \href{../theories/html/hydras.Epsilon0.F_alpha.html}{Epsilon0.F\_alpha}, we prove a relation between the $F$ and $H'$ functional. For any $\alpha$ and $k>0$,
$H'_{\omega^\alpha}(k)\geq F_\alpha(k)$.

\begin{Coqsrc}
H'_F : forall (alpha : E0) (n : nat), 
     F_ alpha (S n) <= H'_ (Phi0 alpha) (S n) 
\end{Coqsrc}

Our proof of this lemma is not trivial at all. One of it sub-goals is the following one:

\begin{Coqsrc}
  alpha : E0
  IHalpha : forall beta : E0, beta o< alpha -> P beta
  Halpha : Limitb alpha
  n : nat
  ============================
  H'_ (Phi0 (CanonS alpha n)) (S n) <= 
  H'_ (Phi0 (CanonS alpha (S n))) (S n)
\end{Coqsrc}

In mathematical notation: $H'_{\omega^{\canonseq{\alpha}{n}}}(n+1) \leq
H'_{\omega^{\canonseq{\alpha}{n+1}}}(n+1)$.

But there exists no lemma saying that, if 
$\beta\leq \alpha$, then $H'_\beta(k)\leq H'_\alpha(k)$, for any $\alpha$ and $\beta$. For instance, 
$H'_{42}(3)=45> H'_\omega(3)=7$.

Looking for lemmas of the form $H'_\beta(k)\leq H'_\alpha(k)$, we find this one (from our library
\href{../theories/html/hydras.Epsilon0.Hprimehtml}{Epsilon0.Hprime}):

\begin{Coqanswer}
H'_restricted_mono_l : 
    forall (alpha beta : E0) (n : nat), 
      Canon_plus n alpha beta -> 
      H'_ beta n <= H'_ alpha n.
\end{Coqanswer}

Thus, it remains to prove that 
there exists a path from ${\omega^{\canonseq{\alpha}{n+1}}}$
to ${\omega^{\canonseq{\alpha}{n}}}$ composed of 
$n+1$-steps.

Fortunately, the Ketonen-Solovay machinery contains three lemmas which help us to build such a path.

\begin{Coqanswer}
KS_thm_2_4_lemma5 :
  forall [i : nat] [alpha beta : T1],
  const_pathS i alpha beta ->
  nf alpha -> alpha <> zero -> 
  const_pathS i (phi0 alpha) (phi0 beta)

KS_thm_2_4 :
  forall [lambda : T1], nf lambda ->limitb lambda ->
  forall i j : nat, i < j -> 
   const_pathS 0 (canonS lambda j) (canonS lambda i)

Cor12_1 :
forall [alpha : T1], nf alpha ->
      forall (beta : T1) (i n : nat),
      beta t1< alpha ->
     i <= n -> const_pathS i alpha beta -> 
     const_pathS n alpha beta
\end{Coqanswer}

\subsubsection{End of the proof}

We finish the proof by comparing several fast growing functions.

\emph{From \href{../theories/html/hydras.Epsilon0.L_alpha.html}{Epsilon0.L\_alpha}}

\begin{Coqsrc}
H'_L_ : forall (alpha : E0) (i : nat), H'_ alpha i <= L_ alpha (S i)
\end{Coqsrc}

@@
